Hilbert space approximate Bayesian Gaussian processes using Stan

### 2020-04-09 Paper/manuscript_v2/manuscript_07_01.tex
edits until mid 5.3

\usepackage{blindtext}
\usepackage{titlesec}
\usepackage[utf8]{inputenc}
\usepackage{times}
\usepackage{microtype}

@@

\usepackage[sort&compress]{natbib}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{float}
\usepackage{soul}
\usepackage{xcolor}
\graphicspath{{./images/}}

@@

\usepackage[titletoc]{appendix}
\usepackage{booktabs}
\usepackage{boldline}
\usepackage{colortbl}

@@

\bibliographystyle{authordate1}
% TODO macro
\newcommand{\todo}[1]{\textcolor{blue}{\textbf{[#1]}}}
\newcommand{\resp}[1]{\textcolor{red}{\textbf{[#1]}}}

@@

\begin{abstract}
Gaussian processes are powerful non-parametric probabilistic models for stochastic functions, however they entail a complexity that is computationally intractable when the number of observations is large, especially when estimated with fully Bayesian methods such as Markov chain Monte Carlo. In this paper, we focus on a novel approach for low-rank approximate Bayesian Gaussian processes, based on a basis functions approximation via Laplace eigenfunctions for stationary covariance functions. The main contribution of this paper is a detailed analysis of the performance and practical implementation of the method in relation to key factors such as the number of basis functions, support of the prediction space, and smoothness of the latent function. We provide intuitive visualizations and recommendations for choosing the values of these factors, which make it easier for users to improve approximation accuracy and computational performance. We also propose a diagnostics for checking that the number of basis functions and the support of the prediction space are adequate given the data. The proposed approach is simple and exhibits an attractive computational complexity due to its linear structure, and it is easy to implement in probabilistic programming frameworks. Several illustrative examples of the performance and applicability of the method in the probabilistic programming language Stan are presented together with the underlying Stan model code.
\end{abstract}

@@

In this paper we propose and evaluate a new framework for fast and accurate inference for fully Bayesian GPs using basis function approximations. We focus on the basis function approximation via Laplace eigenfunctions for stationary covariance functions proposed by \citet{solin2018hilbert}. Using a basis function expansion, a GP is approximated with a linear model which makes inference considerably faster. The linear model structure makes GPs easy to implement as a building block in more complicated models in modular probabilistic programming frameworks, where there is a big benefit if the approximation specific computation is simple. Furthermore, a linear representation of a GP makes it easier to be used as latent function
in non-Gaussian observational models allowing for more modelling flexibility. The basis function approximation via Laplace eigenfunctions can be made arbitrary accurate and the trade-off between computational complexity and approximation accuracy can easily be controlled.

The Laplace eigenfunctions can be computed analytically and they are independent of the particular choice of the covariance function including the hyperparameters. While the pre-computation cost of the basis functions is $O(m^2n)$, the computational cost of learning the covariance function parameters is $O(mn+m)$ in every step of the optimizer or sampler. This is a big advantage in terms of speed for iterative algorithms such as Markov chain Monte Carlo (MCMC). Another advantage is the reduced memory requirements of automatic differentiation methods used in modern probabilistic programming frameworks, such as Stan \citep{carpenter2017stan}, WinBUGS \citep{lunn2000winbugs} and others. This is because the memory requirements of automatic differentiation rather scale with the computational complexity instead of with the usual memory requirements for the posterior density computation \todo{I don't understand the sentence about autodiff and computational complexity} \resp{Res: This sentence escapes me. I think this was written by Aki}. The basis function approach also provides an easy way to apply a non-centered parameterization of GPs, which reduces the posterior dependency between parameters representing the estimated function and the hyperparameters of the covariance function, which further improves MCMC efficiency.

@@

While \citet{solin2018hilbert} have fully developed the mathematical theory behind this specific approximation of GPs, further work is needed for its practical implementation in probabilistic programming frameworks. In this paper, the interactions among the key factors of the method such as the number of basis functions, desired prediction space, or properties of the true functional relationship between covariates and response variable, are investigated and analyzed in detail in relation to the performance and accuracy of the method. Practical recommendations are given for the values of the key factors based on intuitive graphical summaries that encode the recognized relationships. Our recommendations will help users to choose valid and optimized values for these factors, improving performance, and saving time of computation. A diagnosis of whether the chosen values for the number of basis functions and the desired prediction space are adequate to fit to the actual data is proposed.

@@

The remainder of the paper is structured as follows. In Section \ref{ch4_gp}, we introduce GPs, covariance functions and spectral density functions. In Section \ref{ch5_sec_method}, the reduced rank approximation to GPs proposed by \citet{solin2018hilbert} is described. In Section \ref{ch5_sec_accuracy}, the accuracy of these approximations under several conditions using analytical and numerical methods is analyzed. Several case studies in which we fit
exact and approximate GPs to real and simulated data are provided in Section \ref{ch5_sec_cases}. A brief conclusion of the work is made in Section \ref{ch5_sec_conclusion}. Appendix \ref{ch5_app_approx_covfun} includes a brief presentation of the mathematical details behind this Hilbert space approximation of a stationary covariance function, and Appendix \ref{ch5_sec_periodic} presents a low-rank representation of a GP for the particular case of a periodic covariance function. 
Online supplemental material with more case studies illustrating the performance and applicability of the method can be found at \url{https://github.com/gabriuma/basis_functions_approach_to_GP} in the subfolder \url{Paper/online_supplemental_material}.

@@

A GP is a stochastic process which defines the distribution over a collection of random variables indexed by a continuous variable, i.e. $\left\lbrace f(t): t \in \mathcal{T}\right\rbrace$ for some index set $\mathcal{T}$. GPs have the defining property that the marginal distribution of any finite subset of random variables, $\left\lbrace f(t_1), f(t_2), \hdots, f(t_K) \right\rbrace$, is a multivariate Gaussian distribution.

In this work, GPs will take the role of a prior distribution over function spaces for non-parametric latent functions in a Bayesian setting. 
Consider a data set $\mathcal{D} = \left\lbrace \bm{x}_n, y_n \right\rbrace_{n=1}^N$, where $y_n$ is modelled conditionally as $p(y_n|f(\bm{x}_n),\phi)$, where $p$ is some parametric distribution with parameters  $\phi$, and $f$ is an unknown function with GP prior, which depends on an input $\bm{x}_n\in {\rm I\!R}^D$. This generalizes readily to more complex models depending on several unknown functions, for example such as $p(y_n|f(\bm{x}_n),g(\bm{x}_n))$ or multilevel models. Our goal is to obtain the posterior distribution for the value of the function $\tilde{f}=f(\tilde{\bm{x}})$  evaluated at a new input point $\tilde{\bm{x}}$.
% That is, we want to obtain the predictive distribution $p(f^*|\mathcal{D})$ of $f^*$ conditioned on the data $\mathcal{D}$.\\

@@

The mean and covariance functions completely characterize the GP prior, and control the a priori behavior of the function $f$. Let $\bm{f}=\left\lbrace f(\bm{x}_n) \right\rbrace_{n=1}^N$, then the resulting prior distribution for $\bm{f}$ is a multivariate Gaussian distribution
$\bm{f} \sim \mathcal{N}(\bm{\mu}, \bm{K})$,
 where $\bm{\mu} = \left\lbrace \mu(\bm{x}_n) \right\rbrace_{n=1}^N$ is the mean and $\bm{K}$ the covariance matrix, where $K_{i,j}=k(\bm{x}_i,\bm{x}_j)$. In the following, we focus on zero-mean Gaussian processes, that is set $\mu(\bm{x}) = 0$. The covariance function $k(\bm{x}, \bm{x}')$ might depend on a set of hyperparameters, $\bm{\theta}$, but we will not write this dependency explicitly to ease the notation.
The joint distribution of $\bm{f}$ and a new $\tilde{f}$ is also a multivariate Gaussian as,
%
\begin{align*}
p(\bm{f}, \tilde{f})=\mathcal{N} \left( \left[ \begin{array}{cc}
\bm{f} \\ 
f^*
\end{array} \right] \,\middle|\, \bm{0},\left[ \begin{array}{cc}
\bm{K}_{\bm{f},\bm{f}} & \bm{k}_{\bm{f},\tilde{f}} \\ 
\bm{k}_{\tilde{f},\bm{f}} & k_{\tilde{f},\tilde{f}}
\end{array} \right] \right),
\end{align*}

\noindent where $\bm{k}_{\bm{f},\tilde{f}}$ is the covariance between $\bm{f}$ and $\tilde{f}$, and $k_{\tilde{f},\tilde{f}}$ is the prior variance of $\tilde{f}$. %By using the conditioning properties of multivariate Gaussian distributions, we can derive the predictive distribution for $f^*$  given $\bm{f}$ analytically,
% \begin{align*}
	% p(f^*| \bm{f}) = \mathcal{N}(f^*| \bm{k}_{f^*,\bm{f}} \bm{K}_{\bm{f},\bm{f}}^{-1} \bm{f},  k_{f^*,f^*}-\bm{k}_{f^*,\bm{f}} \bm{K}_{\bm{f},\bm{f}}^{-1} \bm{k}_{\bm{f},f^*}  )
% \end{align*}

@@

If $p(y_n|f(\bm{x}_n),\phi)=\mathcal{N}(y_n|f(\bm{x}_n),\sigma)$ then $\bm{f}$ can be integrated out analytically (with a computational cost of $O(n^3)$ for exact GPs and $O(nm^2)$ for sparse GPs). If $p(y_n|f(\bm{x}_n),g(\bm{x}_n))=\mathcal{N}(y_n|f(\bm{x}_n),g(\bm{x}_n))$ or $p(y_n|f(\bm{x}_n),\phi)$ is non-Gaussian, the marginalization does not have a closed form solution. Furthermore, if a prior distribution is imposed on $\phi$ and $\bm{\theta}$ to form a joint posterior for $\phi$, $\bm{\theta}$ and $\bm{f}$, approximate inference such as Markov chain Monte Carlo (MCMC) \citep{brooks_2011}, Laplace approximation (\citep{williams1998bayesian,rasmussen2006gaussian}, expectation propagation \citep{minka2001expectation}, or variational Bayes methods \citep{gibbs2000variational,csato2000efficient} are required. In this paper, we focus on the use of MCMC for integrating over the joint
posterior. MCMC is usually not the fastest approach, but it is flexible and allows
accurate inference and uncertainty estimates for general models in probabilistic programming
settings. We consider the computational costs of GPs specifically from
this point of view.

@@

The covariance function is the crucial ingredient in a GP as it encodes our prior assumptions about the function, and defines a correlation structure which characterizes the correlations between function values at different inputs. A covariance function needs be symmetric and positive semi-definite \citep{rasmussen2006gaussian}. A stationary covariance function is a function of $\bm{\tau}=\bm{x}-\bm{x}' \in {\rm I\!R}^D$, such that it can be written $k(\bm{x},\bm{x}') = k(\bm{\tau})$, which means that the covariance is invariant to translations. Isotropic covariance functions are only depends on the input points through the norm of the difference, $k(\bm{x},\bm{x}') = k(|\bm{x}-\bm{x}'|) = k(r), r\in {\rm I\!R}$, which means that the covariance is both translation and rotation invariant. The most commonly used distance between observations is the L2-norm $(|\bm{x}-\bm{x}'|_{L2})$, also known as Euclidean distance, although other types of distances can be considered.

The Mat\'ern class of isotropic covariance functions is given by, 
%
\begin{align*}
k_{\nu}(r)&=\alpha \, \frac{2^{1-\nu}}{\Gamma(\nu)}\Big(\frac{\sqrt{2\nu}r}{\ell}\Big)^{\!\nu} \! K_{\nu} \Big(\frac{\sqrt{2\nu}r}{\ell}\Big),
\end{align*}

\noindent where $\nu > 0$ is the order the kernel, $K_{\nu}$ the modified Bessel function of the second kind, and the $\ell > 0$ and $\alpha > 0$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular case where $\nu=\infty$ and $\nu=3/2$ are probably the most commonly used kernels \citep{rasmussen2006gaussian}, 
%
\begin{align*}
k_{\infty}(r)&=\alpha \exp\!\! \big(\! -\frac{1}{2} \frac{r^2}{\ell^2}\big),  \\
k_{\frac{3}{2}}(r)&=\alpha\Big(1+\frac{\sqrt{3}r}{\ell}\Big) \! \exp\!\! \big(\! -\frac{\sqrt{3}r}{\ell}\big). 
\end{align*}

\noindent The former is commonly known as the squared exponential (exponentiated quadratic) covariance function. Assuming the Euclidean distance between observations, $r=|\bm{x}-\bm{x}'|_{L2}=\sqrt{\sum_{i=1}^{D}(x_i-x_i')^2}$, the kernels written above take the form
%
\begin{align*}
k_{\infty}(|\bm{x}-\bm{x}'|_{L2})&=\alpha \exp\!\!\Big(\! -\frac{1}{2} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\Big),\\
k_{\frac{3}{2}}(|\bm{x}-\bm{x}'|_{L2})&=\alpha \bigg(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\bigg) \! \exp\!\!\Big(\! - \sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\Big). \\
\end{align*}

\noindent Notice that the previous expressions have been easily generalized to using a multidimensional length-scale $\bm{\ell}\in {\rm I\!R}^D$. Using individual length-scales for each dimension turns the isotropic covariance function into a non-isotropic covariance function. That is, for a non-isotropic covariance function, the smoothness may vary across different input dimensions.

Stationary covariance functions can be represented in terms of their spectral densities \citep{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure \citep[\textit{Bochner's theorem}, see, e.g. ][]{akhiezer1993theory}. If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. The spectral density functions associated with the Mat\'ern class of covariance functions is given by
%
\begin{align*}
S_{\nu}(\omega)&= \alpha \, \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)\, l^{2\nu}}\Big(\frac{2\nu}{l^2}+4\pi^2\omega^2 \Big)^{\!(\nu+D/2)}
\end{align*}

\noindent in $D$ dimensions, where variable $\omega\in {\rm I\!R}$ denotes the frequency, and $\ell$ and $\alpha$ are the lengthscale and magnitude (marginal variance), respectively, of the kernel. The particular cases where $\nu=\infty$ and $\nu=3/2$ take the form
%
\begin{align}
S_{\infty}(\omega)&= \alpha \sqrt{2\pi}^D  \ell^D \exp(-0.5 \ell^2 \omega^2), \label{eq_specdens_inf}  \\
S_{\frac{3}{2}}(\omega)&= \alpha \, \frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2})\sqrt{3}^3}{\frac{1}{2}\sqrt{\pi}\,\ell^3}\Big(\frac{3}{\ell^2}+\omega^2 \Big)^{\!-\frac{D+3}{2}}. \label{eq_specdens_32} 
\end{align}

\noindent For input dimension $D=3$ and Euclidean distance $\omega=\sqrt{\sum_{i=1}^{3}s_i^2}$, and considering a multidimensional lengthscale $\bm{\ell} \in {\rm I\!R}^{3}$, the spectral densities written above take the form
%
\begin{align*}
S_{\infty}(\omega)&= \alpha \sqrt{2\pi}^{3}  \prod_{i=1}^{3} \ell_i  \exp\!\!\big(\!-\frac{1}{2} \sum_{i=1}^{3} \ell_i^2 s_i^2 \big),   \\
S_{\frac{3}{2}}(\omega)&= \alpha \, 32\pi\sqrt{3}^3\prod_{i=1}^{3}\ell_i\big(3+\sum_{i=1}^{3}\ell_i^2 s_i^2 \big)^{\!-3}.
\end{align*}

@@

The approximate GP method, developed by \citet{solin2018hilbert} and implemented in this paper, is based on considering the covariance operator of a stationary covariance function as a pseudo-differential operator constructed as a series of Laplace operators. Then, the pseudo-differential operator is approximated with Hilbert space methods on a compact subset $\Omega \subset {\rm I\!R}^D$ subject to boundary conditions. For brevity, we will refer to these approximate Gaussian processes as HSGPs. Below, we will present the main results around HSGPs relevant for practical applications. More details and mathematical proofs are provided by \citet{solin2018hilbert}. Our starting point for presenting the method is the main result obtained by \citet{solin2018hilbert} of the definition of the covariance function as a series expansion of eigenvalues and eigenfunctions of the Laplacian operator. The mathematical details of this approximation are briefly presented in Appendix \ref{ch5_app_approx_covfun}.

@@

The results from the previous section can be generalizes to a multidimensional input space with compact support, $\Omega=[-L_1,L_1] \times \dots \times [-L_d,L_d]$ and Dirichlet boundary conditions. 
In a $D$-dimensional input space, the total number of eigenfunctions and eigenvalues in the approximation is equal to the number of $D$-tuples, that is possible combinations of univariate eigenfunctions over all dimensions. The number of $D$-tuples is given by 
%
\begin{align} \label{eq_m_multi}
m^{\ast} = \prod_{d=1}^{D} m_d,
\end{align}

>>>>

\usepackage{blindtext}
%\usepackage{titlesec}
\usepackage[utf8]{inputenc}
\usepackage{times}
\usepackage{microtype}

@@

\usepackage{natbib}
\bibliographystyle{plainnat}
\bibpunct{(}{)}{;}{a}{,}{,}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{float}
\usepackage{soul}
\usepackage{xcolor}
\graphicspath{{./images/}}

@@

\usepackage[titletoc]{appendix}
\usepackage{booktabs}
\usepackage{boldline}
\usepackage{colortbl}
\DeclareMathOperator{\LogNormal}{LogNormal}

@@

%\bibliographystyle{authordate1}
% TODO macro
\newcommand{\todo}[1]{\textcolor{blue}{\textbf{[#1]}}}
\newcommand{\resp}[1]{\textcolor{red}{\textbf{[#1]}}}

@@

\begin{abstract}
Gaussian processes are powerful non-parametric probabilistic models for stochastic functions, however they entail a complexity that is computationally intractable when the number of observations is large, especially when estimated with fully Bayesian methods such as Markov chain Monte Carlo. In this paper, we focus on a novel approach for low-rank approximate Bayesian Gaussian processes, based on a basis functions approximation via Laplace eigenfunctions for stationary covariance functions. The main contribution of this paper is a detailed analysis of the performance and practical implementation of the method in relation to key factors such as the number of basis functions, domain of the prediction space, and smoothness of the latent function. We provide intuitive visualizations and recommendations for choosing the values of these factors, which make it easier for users to improve approximation accuracy and computational performance. We also propose a diagnostics for checking that the number of basis functions and the domain of the prediction space are adequate given the data. The proposed approach is simple and exhibits an attractive computational complexity due to its linear structure, and it is easy to implement in probabilistic programming frameworks. Several illustrative examples of the performance and applicability of the method in the probabilistic programming language Stan are presented together with the underlying Stan model code.
\end{abstract}

@@

In this paper, we focus on a new framework for fast and accurate inference for fully Bayesian GPs using basis function approximations based on approximation via Laplace eigenfunctions for stationary covariance functions proposed by \citet{solin2018hilbert}. Using a basis function expansion, a GP is approximated with a linear model which makes inference considerably faster. The linear model structure makes GPs easy to implement as a building block in more complicated models in modular probabilistic programming frameworks, where there is a big benefit if the approximation specific computation is simple. Furthermore, a linear representation of a GP makes it easier to be used as latent function
in non-Gaussian observational models allowing for more modelling flexibility. The basis function approximation via Laplace eigenfunctions can be made arbitrary accurate and the trade-off between computational complexity and approximation accuracy can easily be controlled.

The Laplace eigenfunctions can be computed analytically and they are independent of the particular choice of the covariance function including the hyperparameters. While the pre-computation cost of the basis functions is $O(m^2n)$, the computational cost of learning the covariance function parameters is $O(mn+m)$ in every step of the optimizer or sampler. This is a big advantage in terms of speed for iterative algorithms such as Markov chain Monte Carlo (MCMC). Another advantage is the reduced memory requirements of automatic differentiation methods used in modern probabilistic programming frameworks, such as Stan \citep{carpenter2017stan} and others. This is because the memory requirements of automatic differentiation scale with the size of the autodiff expression tree which in direct implementations is simpler for basis function than covariance matrix based approach. \todo{Better?}. The basis function approach also provides an easy way to apply a non-centered parameterization of GPs, which reduces the posterior dependency between parameters representing the estimated function and the hyperparameters of the covariance function, which further improves MCMC efficiency.

@@

While \citet{solin2018hilbert} have fully developed the mathematical theory behind this specific approximation of GPs, further work is needed for its practical implementation in probabilistic programming frameworks. In this paper, the interactions among the key factors of the method such as the number of basis functions, domain of the prediction space, and properties of the true functional relationship between covariates and response variable, are investigated and analyzed in detail in relation to the computational performance and accuracy of the method. Practical recommendations are given for the values of the key factors based on intuitive graphical summaries that encode the recognized relationships. Our recommendations will help users to choose valid and optimized values for these factors, improving computational performance without sacrificing modeling accuracy. We propose also diagnostics to indicate whether the chosen values for the number of basis functions and the domain of the prediction space are adequate to model the data well.

@@

The remainder of the paper is structured as follows. In Section \ref{ch4_gp}, we introduce GPs, covariance functions and spectral density functions. In Section \ref{ch5_sec_method}, the reduced rank approximation to GPs proposed by \citet{solin2018hilbert} is described. In Section \ref{ch5_sec_accuracy}, the accuracy of these approximations under several conditions using analytical and numerical methods is analyzed. Several case studies in which we fit
exact and approximate GPs to real and simulated data are provided in Section \ref{ch5_sec_cases}. A brief conclusion of the work is made in Section \ref{ch5_sec_conclusion}. Appendix \ref{ch5_app_approx_covfun} includes a brief presentation of the mathematical details behind the Hilbert space approximation of a stationary covariance function, and Appendix \ref{ch5_sec_periodic} presents a low-rank representation of a GP for the particular case of a periodic covariance function. 
Online supplemental material with more case studies illustrating the performance and applicability of the method can be found at \url{https://github.com/gabriuma/basis_functions_approach_to_GP} in the subfolder \url{Paper/online_supplemental_material}.

@@

A GP is a stochastic process which defines the distribution over a collection of random variables indexed by a continuous variable, that is, $\left\lbrace f(t): t \in \mathcal{T}\right\rbrace$ for some index set $\mathcal{T}$. GPs have the defining property that the marginal distribution of any finite subset of random variables, $\left\lbrace f(t_1), f(t_2), \hdots, f(t_K) \right\rbrace$, is a multivariate Gaussian distribution.

In this work, GPs will take the role of a prior distribution over function spaces for non-parametric latent functions in a Bayesian setting. 
Consider a data set $\mathcal{D} = \left\lbrace \bm{x}_n, y_n \right\rbrace_{n=1}^N$, where $y_n$ is modelled conditionally as $p(y_n \mid f(\bm{x}_n),\phi)$, where $p$ is some parametric distribution with parameters  $\phi$, and $f$ is an unknown function with GP prior, which depends on an input $\bm{x}_n\in {\rm I\!R}^D$. This generalizes readily to more complex models depending on several unknown functions, for example such as $p(y_n \mid f(\bm{x}_n),g(\bm{x}_n))$ or multilevel models. Our goal is to obtain the posterior distribution for the value of the function $\tilde{f}=f(\tilde{\bm{x}})$  evaluated at a new input point $\tilde{\bm{x}}$.
% That is, we want to obtain the predictive distribution $p(f^* \mid \mathcal{D})$ of $f^*$ conditioned on the data $\mathcal{D}$.\\

@@

The mean and covariance functions completely characterize the GP prior, and control the a priori behavior of the function $f$. Let $\bm{f}=\left\lbrace f(\bm{x}_n) \right\rbrace_{n=1}^N$, then the resulting prior distribution for $\bm{f}$ is a multivariate Gaussian distribution
$\bm{f} \sim \mathcal{N}(\bm{\mu}, \bm{K})$,
 where $\bm{\mu} = \left\lbrace \mu(\bm{x}_n) \right\rbrace_{n=1}^N$ is the mean and $\bm{K}$ the covariance matrix, where $K_{i,j}=k(\bm{x}_i,\bm{x}_j)$. In the following, we focus on zero-mean Gaussian processes, that is set $\mu(\bm{x}) = 0$. The covariance function $k(\bm{x}, \bm{x}')$ might depend on a set of hyperparameters, $\bm{\theta}$, but we will not write this dependency explicitly to ease the notation.
The joint distribution of $\bm{f}$ and a new $\tilde{f}$ is also a multivariate Gaussian as,
%
\begin{align*}
p(\bm{f}, \tilde{f})=\mathcal{N} \left( \left[ \begin{array}{cc}
\bm{f} \\ 
f^*
\end{array} \right] \,\middle|\, \bm{0},\left[ \begin{array}{cc}
\bm{K}_{\bm{f},\bm{f}} & \bm{k}_{\bm{f},\tilde{f}} \\ 
\bm{k}_{\tilde{f},\bm{f}} & k_{\tilde{f},\tilde{f}}
\end{array} \right] \right),
\end{align*} 
where $\bm{k}_{\bm{f},\tilde{f}}$ is the covariance between $\bm{f}$ and $\tilde{f}$, and $k_{\tilde{f},\tilde{f}}$ is the prior variance of $\tilde{f}$. %By using the conditioning properties of multivariate Gaussian distributions, we can derive the predictive distribution for $f^*$  given $\bm{f}$ analytically,
% \begin{align*}
	% p(f^*| \bm{f}) = \mathcal{N}(f^*| \bm{k}_{f^*,\bm{f}} \bm{K}_{\bm{f},\bm{f}}^{-1} \bm{f},  k_{f^*,f^*}-\bm{k}_{f^*,\bm{f}} \bm{K}_{\bm{f},\bm{f}}^{-1} \bm{k}_{\bm{f},f^*}  )
% \end{align*}

@@

If $p(y_n \mid f(\bm{x}_n),\phi)=\mathcal{N}(y_n \mid f(\bm{x}_n),\sigma)$ then $\bm{f}$ can be integrated out analytically (with a computational cost of $O(n^3)$ for exact GPs and $O(nm^2)$ for sparse GPs). If $p(y_n \mid f(\bm{x}_n),g(\bm{x}_n))=\mathcal{N}(y_n \mid f(\bm{x}_n),g(\bm{x}_n))$ or $p(y_n \mid f(\bm{x}_n),\phi)$ is non-Gaussian, the marginalization does not have a closed form solution. Furthermore, if a prior distribution is imposed on $\phi$ and $\bm{\theta}$ to form a joint posterior for $\phi$, $\bm{\theta}$ and $\bm{f}$, approximate inference such as Markov chain Monte Carlo \citep[MCMC; ][]{brooks_2011}, Laplace approximation \citep{williams1998bayesian,rasmussen2006gaussian}, expectation propagation \citep{minka2001expectation}, or variational inference methods \citep{gibbs2000variational,csato2000efficient} are required. In this paper, we focus on the use of MCMC for integrating over the joint
posterior. MCMC is usually not the fastest approach, but it is flexible and allows
accurate inference and uncertainty estimates for general models in probabilistic programming
settings. We consider the computational costs of GPs specifically from
this point of view.

@@

The covariance function is the crucial ingredient in a GP as it encodes our prior assumptions about the function, and defines a correlation structure which characterizes the correlations between function values at different inputs. A covariance function needs to be symmetric and positive semi-definite \citep{rasmussen2006gaussian}. A stationary covariance function is a function of $\bm{\tau}=\bm{x}-\bm{x}' \in {\rm I\!R}^D$, such that it can be written $k(\bm{x},\bm{x}') = k(\bm{\tau})$, which means that the covariance is invariant to translations. Isotropic covariance functions depend only on the input points through the norm of the difference, $k(\bm{x},\bm{x}') = k(|\bm{x}-\bm{x}'|) = k(r), r\in {\rm I\!R}$, which means that the covariance is both translation and rotation invariant. The most commonly used distance between observations is the L2-norm $(|\bm{x}-\bm{x}'|_{L2})$, also known as Euclidean distance, although other types of distances can be considered.

The Mat\'ern class of isotropic covariance functions is given by, 
%
\begin{align*}
k_{\nu}(r)&=\alpha \, \frac{2^{1-\nu}}{\Gamma(\nu)}\left(\frac{\sqrt{2\nu}r}{\ell}\right)^{\!\nu} \! K_{\nu} \left(\frac{\sqrt{2\nu}r}{\ell}\right),
\end{align*}
where $\nu > 0$ is the order the kernel, $K_{\nu}$ the modified Bessel function of the second kind, and the $\ell > 0$ and $\alpha > 0$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular case where $\nu=\infty$ and $\nu=3/2$ are probably the most commonly used kernels \citep{rasmussen2006gaussian}, 
%
\begin{align*}
k_{\infty}(r)&=\alpha \exp \left( -\frac{1}{2} \frac{r^2}{\ell^2}\right),  \\
k_{\frac{3}{2}}(r)&=\alpha\left(1+\frac{\sqrt{3}r}{\ell}\right) \exp\! \left( -\frac{\sqrt{3}r}{\ell}\right). 
\end{align*}
The former is commonly known as the squared exponential or exponentiated quadratic covariance function. Assuming the Euclidean distance between observations, $r=|\bm{x}-\bm{x}'|_{L2}=\sqrt{\sum_{i=1}^{D}(x_i-x_i')^2}$, the kernels written above take the form
%
\begin{align*}
k_{\infty}(|\bm{x}-\bm{x}'|_{L2})&=\alpha \exp \left( -\frac{1}{2} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\right),\\
k_{\frac{3}{2}}(|\bm{x}-\bm{x}'|_{L2})&=\alpha \left(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\right) \exp\! \left( - \sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\right). \\
\end{align*}
The previous expressions can be easily generalized to using a multidimensional length-scale $\bm{\ell}\in {\rm I\!R}^D$. Using individual length-scales for each dimension turns the isotropic covariance function into a non-isotropic covariance function. That is, for a non-isotropic covariance function, the smoothness may vary across different input dimensions.

Stationary covariance functions can be represented in terms of their spectral densities \citep{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure \citep[\textit{Bochner's theorem}; see, e.g. ][]{akhiezer1993theory}. If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. The spectral density functions associated with the Mat\'ern class of covariance functions are given by
%
\begin{align*}
S_{\nu}(\omega)&= \alpha \, \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)\, l^{2\nu}}\left(\frac{2\nu}{l^2}+4\pi^2\omega^2 \right)^{(\nu+D/2)}
\end{align*}
in $D$ dimensions, where variable $\omega\in {\rm I\!R}$ denotes the frequency, and $\ell$ and $\alpha$ are the lengthscale and magnitude (marginal variance), respectively, of the kernel. The particular cases where $\nu=\infty$ and $\nu=3/2$ take the form
%
\begin{align}
S_{\infty}(\omega)&= \alpha \sqrt{2\pi}^D  \ell^D \exp(-0.5 \ell^2 \omega^2), \label{eq_specdens_inf}  \\
S_{\frac{3}{2}}(\omega)&= \alpha  \frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2})\sqrt{3}^3}{\frac{1}{2}\sqrt{\pi}\ell^3}\left(\frac{3}{\ell^2}+\omega^2 \right)^{-\frac{D+3}{2}}. \label{eq_specdens_32} 
\end{align}
For input dimension $D=3$ and Euclidean distance $\omega=\sqrt{\sum_{i=1}^{3}s_i^2}$, and considering a multidimensional lengthscale $\bm{\ell} \in {\rm I\!R}^{3}$, the spectral densities written above take the form
%
\begin{align*}
S_{\infty}(\omega)&= \alpha \sqrt{2\pi}^{3}  \prod_{i=1}^{3} \ell_i  \exp\!\left(-\frac{1}{2} \sum_{i=1}^{3} \ell_i^2 s_i^2 \right),   \\
S_{\frac{3}{2}}(\omega)&= \alpha \, 32\,\pi\sqrt{3}^3\prod_{i=1}^{3}\ell_i\left(3+\sum_{i=1}^{3}\ell_i^2 s_i^2 \right)^{-3}.
\end{align*}

@@

The approximate GP method, developed by \citet{solin2018hilbert} and further analysed in this paper, is based on considering the covariance operator of a stationary covariance function as a pseudo-differential operator constructed as a series of Laplace operators. Then, the pseudo-differential operator is approximated with \emph{Hilbert space} methods on a compact subset $\Omega \subset {\rm I\!R}^D$ subject to boundary conditions. For brevity, we will refer to these approximate Gaussian processes as HSGPs. Below, we will present the main results around HSGPs relevant for practical applications. More details and mathematical proofs are provided by \citet{solin2018hilbert}. Our starting point for presenting the method is the definition of the covariance function as a series expansion of eigenvalues and eigenfunctions of the Laplacian operator. The mathematical details of this approximation are briefly presented in Appendix \ref{ch5_app_approx_covfun}.

\subsection{Unidimensional GPs} \label{ch5_sec_method_uni}

@@

The results from the previous section can be generalized to a multidimensional input space with compact support, $\Omega=[-L_1,L_1] \times \dots \times [-L_d,L_d]$ and Dirichlet boundary conditions. 
In a $D$-dimensional input space, the total number of eigenfunctions and eigenvalues in the approximation is equal to the number of $D$-tuples, that is, possible combinations of univariate eigenfunctions over all dimensions. The number of $D$-tuples is given by 
%
\begin{align} \label{eq_m_multi}
m^{\ast} = \prod_{d=1}^{D} m_d,
\end{align}
where $m_d$ is the number of basis function for the dimension $d$. Let $\mathbb{S}\in {\rm I\!N}^{m^{\ast} \times D}$ be the matrix of all those $D$-tuples. For example, suppose we have $D=3$ dimensions and use $m_{1}=2$, $m_{2}=2$ and $m_{3}=3$ eigenfunctions and eigenvalues for the first, second and third dimension, respectively. Then, the number of multivariate eigenfunctions and eigenvalues is $m^{\ast} = m_{1} \cdot m_{2} \cdot m_{3} = 12$ and the matrix $\mathbb{S}\in {\rm I\!N}^{12 \times 3}$ is given by
\begin{align}\small
\mathbb{S}=
\left[ {\begin{array}{ccc}
1 & 1 & 1 \nonumber \\
1 & 1 & 2 \\
1 & 1 & 3 \\
1 & 2 & 1 \\
1 & 2 & 2 \\
1 & 2 & 3 \\
2 & 1 & 1 \\
2 & 1 & 2 \\
2 & 1 & 3 \\
2 & 2 & 1 \\
2 & 2 & 2 \\
2 & 2 & 3 
\end{array} } \right].
\end{align}
